Getting and Cleaning Data, Cousera
========================================================

26/04/2014
Sérgio Casca,

# Assignment

In order to read the data I've created a read function. It reads the first 30 rows to determine the dataset attribute types and afterwards reads the whole dataset with the class types already attribued.
```{r}
## source('Scripts/read_data.R')

read_data <- function(directory,filename,...){
    
    filepath <- paste(directory,filename,sep="/")

    temp_ds <- read.table(filepath,nrows = 30)
    classes <- lapply(temp_ds,class)
    
    ds <- read.table(filepath,colClasses=classes,...)
    
    ds
}

```

## 1. Merges the training and the test sets to create one data set.

The first thing I've did was to gather the attributes descriptions. Since the data is all 'character' type, my read_data function always add a column with numeric indexes. So, I'm interested in the second column, with the descriptions. 
The read_data function also attribues 'factor' type to the descriptions so in order to later used them as attributes names I had to cast them to character. 
(Yes, I could have forced it during the read function)
```{r}
# get attributes description data
features <-  read_data('Datasets','features.txt')
features_names <- as.character(features[,2])
```

Then I've loaded the train datasets
```{r}
#load train data
x_train <- read_data('Datasets/train','X_train.txt')
y_train <- read_data('Datasets/train','y_train.txt')
subject_train <- read_data('Datasets/train','subject_train.txt')
```

And I've changed the data.frames colum names by assigning the features descriptions to the x_train.txt dataset, and personal names to the activity indexes and subject indexes datasets.
```{r}
# Adjust column names
names(x_train) <- features_names
names(y_train) <- 'label_idx'
names(subject_train) <- 'subjectIdx'
```

Finally I've binded them all, by column, to a single dataset.
```{r}
# col bind train dataset
train_dataset <- cbind(subject_train,x_train,y_train)
```

Afterwards, I've done the same procedure to the test datasets.
```{r}
#load test data
x_test <- read_data('Datasets/test','X_test.txt')
y_test <- read_data('Datasets/test','y_test.txt')
subject_test <- read_data('Datasets/test','subject_test.txt')

# Adjust column names
names(x_test) <- features_names
names(y_test) <- 'label_idx'
names(subject_test) <- 'subjectIdx'

# col bind test dataset
test_dataset <- cbind(subject_test,x_test,y_test)
```


And finally, I've binded by row both datasets.
```{r}
# create complete training and test datasets
dataset <- rbind(train_dataset,test_dataset)

# remove variables
rm(features,features_names,x_train,y_train,subject_train,train_dataset,x_test,y_test,subject_test,test_dataset)
```

And the original dataset
```{r}
str(dataset)
```

And the last attribute is the label index
(I've subseted the last 4 columns)
```{r}
# last index
last_columns_index <- length(names(dataset))

str(dataset[,(last_columns_index-4):last_columns_index])
```

## 2.Extracts only the measurements on the mean and standard deviation for each measurement. 

In this step I've just selected the column index where the 'std()' and 'mean()' were substrings of the attributes' names. After gathering the indexes, I've just subset the dataset by the gathered column indexes, the first column which was my Subject Index attribue and the last column wich was my ACtivity Index attribute.
By choosing **fixed=TRUE** in **grep** I've leaved out the 'meanFreq()' substrings.
```{r}
# gather indexes of desired attributes
std_columns <- grep("std()",names(dataset),fixed=TRUE)
mean_columns <- grep("mean()",names(dataset),fixed=TRUE)

# subset original dataset
ds <- dataset[,c(1,std_columns,mean_columns,length(names(dataset)))]

# remove variables
rm(dataset,std_columns,mean_columns)
```

And the subset dataset
```{r}
str(ds)
```

## 3. Uses descriptive activity names to name the activities in the data set
## 4. Appropriately labels the data set with descriptive activity names. 

I've joinned steps 3 and 4 in the same piece of code.
I've read the activity labels dataset and renamed the columns names. The activity index column I've nammed the same as the in the original dataset in order to ease the merge operation. 
```{r}
activity_labels <-  read_data('Datasets','activity_labels.txt')
names(activity_labels) <- c('label_idx','activity')
```

Merged the original dataset with the activities descriptions dataset are removed the acitivy indexes column. 
Since the merge operation was by the activity index, this attribute will be the first column on the merged dataset.
```{r}
# merge activity descriptive labels with corresponding index 'activityIndex'
ds_with_labels <- merge(ds[,],activity_labels,by='label_idx')
# let go label_idx
ds_with_labels <- ds_with_labels[,-1]

# remove variables
rm(ds,activity_labels)
```

And the merged dataset
```{r}
str(ds_with_labels)
```

## 5. Creates a second, independent tidy data set with the average of each variable for each activity and each subject. 

In this step, I've just aggregate the data, by the two desired factors, Subject index and Activity, and apply the mean function to all the remaining columns (std and mean attributes).   
```{r}
last_measure <- length(names(ds_with_labels)) - 1
first_measure <- 2

tidy_dataset <- aggregate(ds_with_labels[,first_measure:last_measure],by=list(subjectIdx=ds_with_labels$subjectIdx,activity=ds_with_labels$activity),mean)

# remove variables
rm(ds_with_labels,first_measure,last_measure)
```

Save the data to a text file
```{r}
#save file
write.table(tidy_dataset,'Datasets/tidy_dataset.txt',sep='\t',col.names=TRUE,row.names=FALSE)
```

And finally, return the dataset.
```{r}
# tidy_dataset
str(tidy_dataset)
```
